Supplement - Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-30 submission/supplement.Rmd
make supplementary figure file

---
title:  Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 1in
font-size: 11pt
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \usepackage{siunitx}
   - \spacing{1.5}
   - \pagenumbering{arabic}
   - \usepackage{indentfirst}

bibliography: references.bib
csl: cell-reports.csl
indent: true
editor_options:
  chunk_output_type: console
---
```{r, knitrcommas, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#adds commas to large numbers
library(huxtable)

knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})

```
Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase.

# Abstract

Mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). Loss of repressive chromatin regulators can lead to the aberrant transcription of tissue-specific genes outside of their intended context, however the mechanisms and consequences of their dysregulation are largely unknown. Here, we explored this cellular identity crisis in NDDs by investigating lysine demethylase 5c (KDM5C), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3), in tissue identity. We found male _Kdm5c_ knockout (-KO) mice, which recapitulate key behavioral phenotypes of Claes-Jensen X-linked intellectual disability, aberrantly expresses many liver, muscle, ovary, and testis genes within the amygdala and hippocampus. Gonad-enriched genes expressed in the _Kdm5c_-KO brain were typically unique to germ cells, indicating an erosion of the soma-germline boundary. Germline genes are usually decommissioned in somatic lineages in the post-implantation epiblast, yet _Kdm5c_-KO epiblast-like cells (EpiLCs) aberrantly expressed key regulators of germline identity and meiosis, including _Dazl_ and _Stra8_. Germline gene suppression is sexually dimorphic, as female EpiLCs required a higher dose of KDM5C to maintain germline gene suppression. Using a curated list of mouse germline-enriched genes, we found KDM5C is selectively recruited to a subset of germline gene promoters that contain CpG islands (CGIs) to facilitate DNA CpG methylation (CpGme) during ESC to EpiLC differentiation. However, late stage spermatogenesis genes devoid of promoter CGIs can also become activated in _Kdm5c_-KO cells via ectopic activation by RFX transcription factors. Thus, distinct suppressive mechanisms are recruited to different germline gene classes and ectopic germline transcriptional programs can mirror germ cell development within somatic tissues.

# Introduction

A single genome holds the instructions to generate the myriad of cell types found within an organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001; @lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017]. Human genetic studies revealed mutations in chromatin regulators are a major cause of neurodevelopmental disorders (NDDs)[@gabrieleChromatinBasisNeurodevelopmental2018] and many studies have identified their importance for regulating brain-specific transcriptional programs. Loss of some chromatin regulators can also result in the ectopic expression of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. However, the mechanisms underlying ectopic gene expression and its impact upon neurodevelopment are poorly understood.

To elucidate the role of tissue identity in chromatin-linked NDDs, it is essential to first characterize the nature of the dysregulated genes and the molecular mechanisms governing their de-repression. Here we focus on lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), which erases histone 3 lysine 4 di- and trimethylation (H3K4me2/3), a permissive chromatin modification enriched at gene promoters[@iwaseXLinkedMentalRetardation2007]. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534). MRXSCJ is more common and severe in males and its neurological phenotypes include intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Male _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ phenotypes, including hyperaggression, increased seizure propensity, and learning impairments[@iwaseMouseModelXlinked2016; @scandagliaLossKdm5cCauses2017]. RNA sequencing (RNA-seq) of the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of some germline genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is unclear if other tissue-specific genes are aberrantly transcribed with KDM5C loss, at what point in development germline gene misexpression begins, and what mechanisms underlie their dysregulation.

Distinguishing between germ cells and somatic cells is a key feature of multicellularity[@devlinPanmetazoanViewGermlinesoma2023] that occurs during early embryogenesis in many metazoans[@lehmannGermlineStemCells2012]. In mammals, chromatin regulators are crucial for decommissioning germline genes during the transition from na&#239;ve to primed pluripotency. Initially, germline gene promoters gain repressive histone H2A lysine 119 monoubiquitination (H2AK119ub1)[@endohPCGF6PRC1SuppressesPremature2017] and histone 3 lysine 9 trimethylation (H3K9me3)[@endohPCGF6PRC1SuppressesPremature2017; @mochizukiRepressionGermlineGenes2021] in embryonic stem cells (ESCs) and are then decorated with DNA CpG methylation (CpGme) in the post-implantation embryo[@velascoDnmt3bRecruitmentE2F62010; @hackettPromoterDNAMethylation2012; @mochizukiRepressionGermlineGenes2021]. The contribution of KDM5C to this process remains unclear. Furthermore, studies on germline gene repression have primarily been conducted in males and focused on marker genes important for germ cell development rather than germline genes as a whole, given the lack of a curated list for germline-enriched genes. Therefore, it is unknown if the mechanism of repression differs between sexes or for certain classes of germline genes, e.g. meiotic genes versus spermatid differentiation genes.

To illuminate KDM5C's role in tissue identity, here we characterized the aberrant expression of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like stem cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We curated list of mouse germline-enriched genes, which enabled genome-wide analysis of germline gene silencing mechanisms for the first time. Based on the data presented below, we propose KDM5C plays a fundamental, sexually dimorphic role in the development of tissue identity during early embryogenesis, including the establishment of the soma-germline boundary.

# Results

## Tissue-enriched genes are aberrantly expressed in the _Kdm5c_-KO brain

```{r, DESeq2_params, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../code/utilities/parameters.R")
```

Previous RNA sequencing (RNA-seq) of the adult male _Kdm5c_-KO hippocampus revealed ectopic expression of some germline genes unique to the testis[@scandagliaLossKdm5cCauses2017]. It is currently unknown if the testis is the only tissue type misexpressed in the _Kdm5c_-KO brain. We thus characterized the role of KDM5C in brain tissue identity by systematically assessing the dysregulation of genes enriched in 17 mouse tissues[@liComprehensiveMouseTranscriptomic2017]. We analyzed tissue-specific differentially expressed genes (DEGs) in our published mRNA-seq datasets[@vallianatosMutuallySuppressiveRoles2020] of the adult amygdala and hippocampus from wild-type and constitutive _Kdm5c_-KO male mice (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco`, q < 0.1).

```{r, tissue_genes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the number and stats for tissue-enriched gene expression
tissue_res <- read.csv("../data/processed/TissueSpecific_AMYHIP_numberofgenesDEGs.csv", header = TRUE)

#get the value you want to print 
#col is which data column, tiss is which tissue
gettissval <- function(col, tiss){
	val <- subset(tissue_res, tissue == tiss)
	val <- val[,col]
	return(val)
}

```

We found a large proportion of significantly upregulated genes within the _Kdm5c_-KO brain are typically enriched within non-brain tissues (Amygdala: 35%, Hippocampus: 24%) (Figure 1A-B). For both the amygdala and hippocampus, the majority of tissue-enriched (DEGs) were testis genes (Figure 1A-C). Even though the testis has the largest total number of tissue-biased genes (`r gettissval("count", "Testis")` genes) compared to any other tissue, testis-biased DEGs were significantly enriched for both brain regions (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Testis"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Testis"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Testis"), 3)`, Fisher's Exact Test). One example of a testis-enriched gene misexpressed in the _Kdm5c_-KO brain is _FK506 binding protein 6 (Fkbp6)_, a known regulator of PIWI-interacting RNAs (piRNAs) and meiosis[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012] (Figure 1C).

Interestingly, we also observed significant enrichment of ovary-biased DEGs in both the amygdala and hippocampus (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Ovary"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Ovary"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Ovary"), 3)`, Fisher's Exact) (Figure 1A-B). Ovary-enriched DEGs included _Zygotic arrest 1 (Zar1)_, which sequesters mRNAs in oocytes for meiotic maturation[@chengMammalianOocytesStore2022] (Figure 1D). Given that the _Kdm5c_-KO mice we analyzed are male, these data demonstrate that the ectopic expression of gonad-enriched genes is independent of organismal sex.

Although not consistent across brain regions, we also found significant enrichment of DEGs biased towards two non-gonadal tissues - the liver (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Liver"), 2)`, Odds Ratio = `r signif(gettissval("Amygdala_oddsratio", "Liver"), 3)`, Fisher's Exact Test) and the muscle (Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Muscle"), 2)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Muscle"), 3)`, Fisher's Exact Test) (Figure 1A-B). _Apolipoprotein C-I (Apoc1)_ a lipoprotein metabolism and transport gene, is among the liver-biased DEG derepressed in both the hippocampus and amygdala[@roulandRoleApolipoproteinC12022] and its brain overexpression has been implicated in Alzheimer's disease[@leducAPOECholesterolHomeostasis2010] (Figure 1E).

For all _Kdm5c_-KO tissue-enriched DEGs, aberrantly expressed mRNAs are polyadenylated and spliced into mature transcripts (Figure 1C-E). Of note, we observed little to no dysregulation of brain-enriched genes (Amygdala p = `r signif(gettissval("Amygdala_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`; Hippocampus p = `r signif(gettissval("Hippocampus_pvalues", "Brain"), 3)`, Odds Ratio = `r signif(gettissval("Hippocampus_oddsratio", "Brain"), 3)`, Fisher's Exact), despite the fact these are brain samples and the brain has the second highest total number of tissue-enriched genes (`r gettissval("count", "Brain")` genes). Altogether, these results suggest the aberrant expression of tissue-enriched genes within the brain is a major effect of KDM5C loss.

```{r, germgenes, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
germgenes <- read.csv("../data/processed/germGENES20.csv", header = TRUE)

#testis gene ontology results
testisGO <- read.csv("../results/GO_Brain_TestisDEGs.csv")

#function to get the GO ID
getGO <- function(x, df, column, cluster){
	df_test <- df

	if (cluster == "no") {
		out <- df_test[df_test$Description == x,][column][,1]
	} else {
		out <- df_test[df_test$Description == x & df_test$Cluster == cluster,][column][,1]
	}
	outformat <- ifelse(column == "ID", toString(out), out) #an attempt to remove the commas from GO ID numbers, didn't work
	return(outformat) 
}

```
### Germline genes are misexpressed in the _Kdm5c_-KO brain

_Kdm5c_-KO brain expresses testicular germline genes[@scandagliaLossKdm5cCauses2017], however the testis also contains somatic cells that support hormone production and germline functions. To determine if _Kdm5c_-KO results in ectopic expression of somatic testicular genes, we first evaluated the known functions of testicular DEGs through gene onotology. We found _Kdm5c_-KO testis-enriched DEGs had high enrichment of germline-relevant ontologies, including spermatid development (GO: 0007286, p.adjust = `r signif(getGO("spermatid development", testisGO, 'p.adjust', "no"), 3)`) and sperm axoneme assembly (GO: 0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", testisGO, 'p.adjust', "no"), 3)`) (Figure 2A).

We then evlauted testicular DEG expression in wild-type testes versus testes with germ cell depletion[@muellerIndependentSpecializationHuman2013], which was accomplished by heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ (Kit^W/Wv^)[@handelSertoliCellDifferentiation1979]. Almost all _Kdm5c_-KO testis-enriched DEGs lost expression with germ cell depletion (Figure 2B). We then assessed testis-enriched DEG expression in a published single cell RNA-seq dataset that identified cell type-specific markers within the testis[@greenComprehensiveRoadmapMurine2018]. Some _Kdm5c_-KO testis-enriched DEGs were classified as specific markers for different germ cell developmental stages (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids), yet none marked somatic cells (Figure 2C). Together, these data demonstrate that the _Kdm5c_-KO brain aberrantly expresses germline genes, but not somatic testicular genes, reflecting an erosion of the soma-germline boundary.

As of yet, research on germline gene silencing mechanisms has focused on a handful of key genes rather than assessing germline gene suppression genome-wide due to the lack of a comprehensive gene list. We therefore generated a list of mouse germline-enriched genes using RNA-seq datasets of Kit^W/Wv^ mice that included males and females at embryonic day 12, 14, and 16[@sohGeneRegulatoryProgram2015] and adult male testes[@muellerIndependentSpecializationHuman2013]. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type gonads 2) their expression in any non-gonadal tissue of adult wild type mice[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of their maximum expression in the wild-type germline, and 3) their expression in the germ cell-depleted gonads, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource to globally characterize germline gene misexpression with _Kdm5c_ loss (Supplementary table 1).

## _Kdm5c_-KO epiblast-like cells aberrantly express key regulators of germline identity

```{r, epilcs, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}

#testis gene ontology results
epigerm <- read.csv("../results/DESeq2/germDEGs/germDEGs_EpiLC_XY5cKO.csv")
GO_EpiBrain <- read.csv("../results/GO_EpiLC_vs_Brain.csv")

epi_restab <- read.csv("../data/processed/restable_EpiLC_XY5cKO.csv", row.names = 1)
epi_restab$ENSEMBL <- rownames(epi_restab)

zscan4d_padj <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['padj'][,1]
duxf3_padj <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['padj'][,1]

zscan4d_l2fc <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000090714',]['log2FoldChange'][,1]
duxf3_l2fc <- epi_restab[epi_restab$ENSEMBL == 'ENSMUSG00000075046',]['log2FoldChange'][,1]

getresval <- function(df, col, symbol){
	val <- subset(df, SYMBOL == symbol)
	val <- val[,col]
	return(val)
}

```

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014], when they are silenced in epiblast stem cells that will  differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of na&#239;ve embryonic stem cells (nESCs) into epiblast-like stem cells (EpiLCs) (Figure 3A)[@hayashiReconstitutionMouseGerm2011; @samantaGeneratingPrimedPluripotent2020]. While some germline-enriched genes are also expressed in nESCs and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@borgelTargetsDynamicsPromoter2010; @mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for the initial silencing germline genes in somatic lineages by evaluating the impact of _Kdm5c_ loss in male EpiLCs.

_Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation appeared normal (Figure 3B) and EpiLCs properly expressed markers of primed pluripotency, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C). We then identified tissue-enriched DEGs in our previously published RNA-seq dataset of wild-type and _Kdm5c_-KO EpiLCs[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1). Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3D). Using our list of mouse germline-enriched genes assembled above, we found `r nrow(epigerm)` germline genes were misexpressed in male _Kdm5c_-KO EpiLCs.

We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if germline genes are constitutively dysregulated or change over the course of development. The majority of germline DEGs were unique to either EpiLCs or the brain, with only _Cyct_ shared across all tissue/cell types (Figure 3E-F). EpiLCs had particularly high enrichment of meiosis-related gene ontologies (Figure 3G), such as meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO:0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes involved in sperm axoneme assembly (GO:0007288, p.adjust = `r signif(getGO("sperm axoneme assembly", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`) and sperm motility (GO:0097722, p.adjust = `r signif(getGO("sperm motility", GO_EpiBrain, 'p.adjust', "Hippocampus"), 3)`).

Notably, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8_ (_Stra8_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Stra8'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Stra8'), 3)`) and _Deleted in azoospermia like_ (_Dazl_: log2 fold change = `r signif(getresval(epigerm, 'log2FoldChange', 'Dazl'), 3)`, q = `r signif(getresval(epigerm, 'padj', 'Dazl'), 3)`) (Figure 3H). These genes are typically expressed when primordial germ cells (PGCs) are committed to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo[@macfarlanEmbryonicStemCell2012; @guptaGenomewideScreenReveals2023]. However, we did not see derepression of two-cell stage-specific genes, like _Duxf3 (Dux)_ (log2 fold change = `r signif(duxf3_l2fc, 3)`, q = `r signif(duxf3_padj, 3)`) and _Zscan4d_ (log2 fold change =  `r signif(zscan4d_l2fc, 3)`, q = `r signif(zscan4d_padj, 3)`) (Figure 3H), indicating _Kdm5c_-KO EpiLCs do not revert back to a 2-cell-like state. Altogether, _Kdm5c_-KO EpiLCs expressing key drivers of germline identity and meiosis while the brain primarily expresses spermiogenesis genes indicate germline gene misexpression parallels germline development during proper _Kdm5c_-KO differentiation.

## Female epiblast-like cells have increased sensitivity to germline gene misexpression with _Kdm5c_ loss
```{r, DESeq2_params_XX, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
#get the parameters used repeatedly, like the log2fold change cut off
source("../../KDM5C_Germ_Dev_Sex_RA/code/utilities/parameters.R")
```

It is currently unknown if the misexpression of germline genes is influenced by sex, as previous studies on germline gene repressors have focused on males[@pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Sex is particularly pertinent in the case of KDM5C because it partially escapes X chromsome inactivation (XCI), resulting in a higher dosage in females[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. We therefore explored the impact of chromosomal sex upon germline gene suppression by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO), and female heterozygous knockout (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

Homozygous and heterozygous _Kdm5c_ knockout females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), many were male-specific and female-specific, such as _Tktl2_ and _Esx1_ (Figure 4B). We then compared the known functions of germline genes dysregulated only in females (XX only - dysregulated in XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), only in males (XY only), or in all samples (shared) (Figure 4C). Female-specific germline DEGs were enriched for meiotic (GO:0051321 meiotic cell cycle) and flagellar (GO:0003341 cilium movement) functions, while male-specifc DEGs had roles in mitochondrial and cell signaling (GO:0070585 protein localization to mitochondrion). Germline transcripts expressed in both sexes were enriched for meiotic (GO:0140013 meiotic nuclear division) and egg-specific functions (GO:0007292 female gamete generation).

The majority of germline genes expressed in both sexes had a greater log2 fold change in females compared to males (Figure 4D-F). This increased degree of dysregulation in females, along with the increased total number of germline genes, indicates females are more sensitive to losing KDM5C-mediated germline gene suppression. Female sensitivity could be due to impaired XCI in _Kdm5c_ mutants[@samantaActivationXistEvolutionarily2022], as many spermatogenesis genes lie on the X chromosome[@wangAbundanceXlinkedGenes2001; @khilMouseChromosomeEnriched2004]. However, female germline DEGs were not biased towards the X chromosome and had a a similar overall proportion of X chromosome DEGs compared to males (XY _Kdm5c_-KO - 10.29%, XX _Kdm5c_-HET - 7.43%, XX _Kdm5c_-KO - 10.59%) (Figure 4G). The majority of germline DEGs instead lie on autosomes for both male and female _Kdm5c_ mutants (Figure 4G). Thus, while female EpiLCs are more prone to germline gene misexpression with KDM5C loss, it is likely independent of XCI defects.

### Germline gene misexpression in _Kdm5c_ mutants is independent of germ cell sex

Although many germline genes have shared functions in the male and female germline, some have unique or sex-biased expression. Therefore, we wondered if _Kdm5c_ mutant males would primarily express sperm genes while mutant females primarily expressed egg genes. To comprehensively assess whether germline gene sex corresponds with _Kdm5c_ mutant sex, we first filtered our list of germline-enriched genes for egg and sperm-biased genes (Figure 4H). We defined germ cell sex-biased genes as those whose expression in the opposite sex, at any time point, is no greater than 20% of the gene's maximum expression in a given sex. This yielded `r nrow(subset(germgenes, sexBias == "XX"))` egg-biased, `r nrow(subset(germgenes, sexBias == "XY"))` sperm-biased, and `r nrow(subset(germgenes, sexBias == "unbiased"))` unbiased germline-enriched genes. We found egg, sperm, and unbiased germline genes were dyregulated in all _Kdm5c_ mutants at similar proportions (Figure 4I-J). Furthermore, germline genes dysregulated exclusively in either male or female mutants were also not biased towards their corresponding germ cell sex (Figure 4I). Altogether, these results demonstrate sex differences in germline gene dysregulation is not due to sex-specific activation of sperm or egg transcriptional programs.

## KDM5C binds to a subset of germline gene promoters during early embryogenesis

KDM5C binds to the promoters of several germline genes in embryonic stem cells (ESCs) but its binding is absent in neurons[@scandagliaLossKdm5cCauses2017]. However, the lack of a comprehensive list of germline-enriched genes prohibited genome-wide characterization of KDM5C binding at germline gene promoters. Thus, it is unclear if KDM5C is enriched at germline gene promoters, what types of germline genes KDM5C regulates, and if its binding is maintained at any germline genes in neurons.

```{r, KDM5C_ChIPseq, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
KDM5C_ChIP_EpiLC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv")

KDM5C_ChIP_PNC_promo <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv")

KDM5C_ChIP_EpiLC <- read.csv("../data/raw/ChIPseq_mm10_EpiLC_WTnoKO_consensus_peaks.bed", sep = "\t")

KDM5C_ChIP_PNC <- read.csv("../data/raw/ChIPseq_mm10_PNC_WTnoKO_consensus_peaks.bed", sep = "\t")

GO_KDM5C_ChIP <- read.csv("../results/ChIPseq_KDM5C_GO_EpiLC_vs_PNC_unique.csv")

```

To address these questions, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary forebrain neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. EpiLCs had a higher total number of high-confidence KDM5C peaks than PNCs (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1). KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC_promo)`, PNCs: `r nrow(KDM5C_ChIP_PNC_promo)` &#177; `r promorange`bp from the TSS), although PNCs showed increased localization to non-promoter regions (Figure 5A).

The majority of promoters bound by KDM5C in PNCs were also bound in EpiLCs (513 shared promoters), however a large portion of gene promoters were only bound by KDM5C in EpiLCs (3,677 EpiLC only promoters) (Figure 5B). Genes bound by KDM5C in both PNCs and EpiLCs were enriched for functions involving nucleic acid turnover, such as deoxyribonucleotide metabolic process (GO:0009262, p.adjust = `r signif(getGO("deoxyribonucleotide metabolic process", GO_KDM5C_ChIP, 'p.adjust', "Shared"), 3)`) (Figure 5C). Germline-specific ontologies were only enriched in KDM5C-bound promoters unique to EpiLCs, such as meiotic nuclear division (GO: 0007127 p.adjust = `r signif(getGO("meiosis I", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) and meiotic cell cycle process (GO:1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_KDM5C_ChIP, 'p.adjust', "EpiLC only"), 3)`) (Figure 5C). There were no ontologies significantly enriched for genes bound by KDM5C only in PNCs. Using our mouse germline gene list, we observed evident KDM5C signal around the TSS of many germline genes in EpiLCs, but not in PNCs (Figure 5D). Based on our ChIP-seq peak cut-off criteria, KDM5C was highly enriched at 211 germline gene promoters in EpiLCs (`r signif(211/1288, 3)`% of all germline genes) (Figure 5E). Of note, KDM5C was only bound to about one third of _Kdm5c_-KO RNA-seq DEG promoters (EpiLC only DEGs: `r 18/50*100`%, Brain only DEGs: `r signif(7/21*100,3)`%) (Supplementary figure 1A-C). However, KDM5C was bound to the promoter at 3 out of the 4 genes dysregulated in both the brain and EpiLCs. Representative examples of KDM5C-bound and unbound germline DEGs are _Dazl_ and _Stra8_, repsectively (Figure 5F). Together, these results demonstrate KDM5C is recruited to a subset of germline genes in EpiLCs, including meiotic genes, but does not directly regulate germline genes in neurons. Furthermore, the majority of germline mRNAs expressed in _Kdm5c_-KO cells are dysregulated independent of direct KDM5C binding to their promoters.

Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. To elucidate if KDM5C is recruited to germline genes by a similar mechanism, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I, Supplementary figure 1D). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E). Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.

## KDM5C is recruited to germline gene promoters harboring CpG islands to facilitate _de novo_ DNA methylation

In the early embryo, germline gene promoters are initially decorated with repressive histone modifications and are then silenced long-term via DNA CpG methylation (CpGme)[@borgelTargetsDynamicsPromoter2010; @endohPCGF6PRC1SuppressesPremature2017; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Our results above indicate KDM5C also acts at germline gene promoters during this time period. However, how KDM5C interacts with other germline gene silencing mechanisms is currently unclear. KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007], yet KDM5C's catalytic activity was recently shown to be dispensible for suppressing _Dazl_ in undifferentiated ESCs[@guptaGenomewideScreenReveals2023]. Since H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015], KDM5C's catalytic activity may instead be required in the post-implantation embryo for long-term silencing of germline genes. In support of this, CpGme is markedly reduced at two germline gene promoters in the _Kdm5c_-KO adult hippocampus[@scandagliaLossKdm5cCauses2017].

>>>>

---
title: Supplement - Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C
output: pdf_document
geometry: margin = 1in
font-size: 11pt
header-includes:
   - \renewcommand{\figurename}{Supplementary Figure}
   - \usepackage{caption}
   - \captionsetup[figure]{font=footnotesize, labelfont=bf}
   - \usepackage{sectsty} 
   - \usepackage{helvet}
   - \renewcommand{\familydefault}{\sfdefault}
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand\linenumberfont{\normalfont\small}
   - \usepackage{setspace}
   - \usepackage{siunitx}
   - \spacing{1.5}
   - \pagenumbering{arabic}
   - \usepackage{indentfirst}

bibliography: references.bib
csl: cell-reports.csl
indent: true
editor_options:
  chunk_output_type: console
---
